---
title: "MLB ABS Challenge Run Value"
subtitle: "Frozen 2026 regular-season snapshot through August 25"
author: "Research pipeline output"
date: last-modified
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(arrow)
library(data.table)
library(ggplot2)
library(knitr)

required_files <- file.path("data", "processed", c(
  "pitch_ledger.parquet", "challenge_events.parquet",
  "remaining_opportunities.parquet", "team_summaries.parquet"
))
if (!all(file.exists(required_files))) {
  stop("Derived files are missing. Run `make pipeline` before rendering.")
}

ledger <- as.data.table(read_parquet(required_files[[1]]))
challenges <- as.data.table(read_parquet(required_files[[2]]))
published_challenges <- challenges[publication_eligible == TRUE]
remaining <- as.data.table(read_parquet(required_files[[3]]))
teams <- as.data.table(read_parquet(required_files[[4]]))
team_intervals <- as.data.table(read_parquet("data/processed/team_intervals.parquet"))
adjusted_intervals <- as.data.table(read_parquet("data/processed/adjusted_intervals.parquet"))
validation <- fread("data/processed/validation_checks.csv")
coverage <- fread("data/processed/coverage.csv")
model_validation <- fread("data/processed/model_validation.csv")
history_coverage <- fread("data/processed/history_coverage.csv")

team_lookup <- unique(rbind(
  ledger[, .(team_id = away_team_id, team = away_team)],
  ledger[, .(team_id = home_team_id, team = home_team)]
))
```

# Provenance and validation

The analysis joins MLB live feeds, daily Statcast pitch exports, and Savant's
ABS detail. Raw responses are cached with SHA-256 manifests. A challenge is
published only when its identity and explicit outcome reconcile to both the
game's final successful/failed totals and Savant.

Statcast alone is not the challenge ledger. Challenge timing and explicit
overturned/upheld outcomes come from MLB's `MJ` review records; Statcast
provides pitch geometry and the pre-pitch state. Initial and corrected
post-call states are generated from that common state, and Savant independently
validates the event identity, original call, outcome, and geometric ruling.
The full acquisition and reconciliation history is recorded in
[`docs/data-acquisition-walkthrough.md`](../docs/data-acquisition-walkthrough.md).

```{r}
kable(validation[, .(check, passed, observed, expected)], digits = 4,
  caption = "Publication acceptance checks")
kable(coverage[scope == "league"], digits = 4)
kable(history_coverage, digits = 4)

model_display <- data.table(
  diagnostic = c("RE288: 2025 holdout", "Home WP: 2025 holdout",
    "RE288 vs Savant overturns"),
  n = model_validation$n,
  metric_1 = c(
    sprintf("RMSE %.4f", model_validation$rmse[[1]]),
    sprintf("Brier %.4f", model_validation$brier[[2]]),
    sprintf("RMSE %.4f", model_validation$rmse[[3]])
  ),
  metric_2 = c(
    sprintf("MAE %.4f", model_validation$mae[[1]]),
    sprintf("log loss %.4f", model_validation$log_loss[[2]]),
    sprintf("correlation %.4f", model_validation$correlation[[3]])
  )
)
kable(model_display, caption = "Model validation diagnostics")
```

# Challenge timing and success-rate accounting

Explicit `false` outcomes are upheld challenges, not missing observations.
The denominator below is therefore every resolved attempt.

```{r}
challenge_accounting <- published_challenges[, .(
  attempts = .N,
  overturned = sum(challenge_outcome == "overturned"),
  upheld = sum(challenge_outcome == "upheld"),
  unresolved = sum(!challenge_outcome %in% c("overturned", "upheld")),
  success_rate = mean(challenge_outcome == "overturned")
)]
kable(challenge_accounting, digits = 4)

timing <- published_challenges[, .N, by = .(inning, half, challenge_outcome)]
ggplot(timing, aes(inning, N, fill = challenge_outcome)) +
  geom_col() +
  facet_wrap(~half) +
  labs(x = "Inning", y = "Challenges", fill = "Outcome") +
  theme_minimal()
```

# Actual challenge value

Immediate gain is the count-aware corrected-state minus original-state value
from the challenger's perspective. Upheld calls have zero immediate gain and
their modeled marginal inventory cost is reported separately.

```{r}
actual <- published_challenges[, .(
  attempts = .N,
  overturned = sum(challenge_outcome == "overturned"),
  upheld = sum(challenge_outcome == "upheld"),
  gross_re = sum(actual_re_gain, na.rm = TRUE),
  failed_inventory_re = sum(failed_inventory_cost_re, na.rm = TRUE),
  net_re = sum(actual_re_gain - failed_inventory_cost_re, na.rm = TRUE),
  gross_wpa = sum(actual_wpa_gain, na.rm = TRUE),
  failed_inventory_wpa = sum(failed_inventory_cost_wpa, na.rm = TRUE)
)][, success_rate := overturned / attempts]
kable(actual[, .(
  attempts, overturned, upheld,
  success_pct = 100 * success_rate,
  gross_re, failed_inventory_re, net_re,
  gross_wpa, failed_inventory_wpa
)], digits = 3, caption = "League challenge value")

failed_cost_benchmark <- published_challenges[challenge_outcome == "upheld", .(
  upheld_attempts = .N,
  savant_fixed_cost_re_per_failure = mean(sz_challenge_lost_runs, na.rm = TRUE),
  modeled_inventory_re_per_failure = mean(failed_inventory_cost_re, na.rm = TRUE),
  modeled_inventory_wpa_per_failure = mean(failed_inventory_cost_wpa, na.rm = TRUE)
)]
setnames(failed_cost_benchmark, c(
  "upheld", "Savant fixed RE", "modeled inventory RE", "modeled inventory WPA"
))
kable(failed_cost_benchmark, digits = 4,
  caption = "Per-failure benchmark versus modeled inventory cost")
```

# Missed and out-of-challenges value

These are realized correctable calls. The out-of-challenges column does not
assume that a club would have challenged every such pitch.

```{r}
missed <- ledger[correctable_opportunity == TRUE, .(
  missed_available = sum(opportunity_status == "missed_available"),
  missed_re = sum(fifelse(opportunity_status == "missed_available",
    potential_challenger_re, 0), na.rm = TRUE),
  out_of_challenges = sum(opportunity_status == "out_of_challenges"),
  out_re = sum(fifelse(opportunity_status == "out_of_challenges",
    potential_challenger_re, 0), na.rm = TRUE)
), by = .(team_id = adverse_team_id)][order(-out_re)]
missed <- merge(missed, team_lookup, by = "team_id")
missed_display <- missed[order(-out_re)][1:10, .(
  Team = team,
  `Missed available` = missed_available,
  `Missed RE` = missed_re,
  `Out of challenges` = out_of_challenges,
  `Out RE` = out_re
)]
kable(missed_display, digits = 3,
  caption = "Ten teams with the most realized RE lost at zero inventory")
```

# Remaining-opportunity inventory curves

Inventory is expressed first in future correctable opportunities, then RE and
WPA. Curves use leave-one-game-out state averages; marginal values use the
observed-behavior recursion.

```{r}
start_states <- remaining[, .SD[which.min(pitch_order)],
  by = .(game_pk, team_id)]
inventory_summary <- start_states[, .(
  expected_correctable = mean(expected_correctable_remaining, na.rm = TRUE),
  expected_offense = mean(expected_correctable_offense_remaining, na.rm = TRUE),
  expected_defense = mean(expected_correctable_defense_remaining, na.rm = TRUE),
  expected_re = mean(expected_re_remaining, na.rm = TRUE),
  expected_wpa = mean(expected_wpa_remaining, na.rm = TRUE),
  first_unit_re = mean(marginal_re_1_to_0, na.rm = TRUE),
  second_unit_re = mean(marginal_re_2_to_1, na.rm = TRUE)
)]
setnames(inventory_summary, c(
  "Correctable", "Offense", "Defense", "Expected RE", "Expected WPA",
  "Unit 1 RE", "Unit 2 RE"
))
kable(inventory_summary, digits = 4,
  caption = "League-average remaining opportunity at game start")

curves <- remaining[, .(
  expected_correctable = mean(expected_correctable_remaining, na.rm = TRUE),
  marginal_first_re = mean(marginal_re_1_to_0, na.rm = TRUE),
  marginal_second_re = mean(marginal_re_2_to_1, na.rm = TRUE)
), by = .(inning, outs_before)]
ggplot(curves, aes(inning, expected_correctable, color = factor(outs_before))) +
  geom_line(linewidth = 0.8) +
  labs(x = "Inning", y = "Expected correctable opportunities remaining",
    color = "Outs") +
  theme_minimal()
```

# Team comparisons and WPA context

Raw totals and opportunity-adjusted attempt residuals are shown together;
neither is interpreted causally.

```{r}
teams_display <- merge(teams, team_lookup, by = "team_id")

decision_table <- teams_display[, .(
  team, attempts, overturns, upheld,
  success_pct = 100 * success_rate,
  net_re, net_wpa
)][order(-net_re)]
kable(decision_table, digits = 3,
  caption = "Challenge outcomes and net value by team")

opportunity_table <- teams_display[, .(
  Team = team,
  `Missed RE` = missed_available_re,
  `Zero-inventory RE` = out_of_challenges_re,
  `Expected left` = expected_correctable_remaining,
  `Unit 1 RE` = marginal_re_first_unit,
  `Unit 2 RE` = marginal_re_second_unit
)][order(-`Zero-inventory RE`)]
kable(opportunity_table, digits = 3,
  caption = "Missed, zero-inventory, and remaining-opportunity value")

adjusted_residual_intervals <- adjusted_intervals[
  metric == "challenge_residual"
]
adjusted_table <- merge(
  teams_display[, .(team_id, team, observed_attempts, expected_attempts,
    challenge_residual)],
  adjusted_residual_intervals[, .(team_id, lower, upper)],
  by = "team_id"
)[order(-challenge_residual)]
kable(adjusted_table[, .(team, observed_attempts, expected_attempts,
  challenge_residual, lower, upper)], digits = 3,
  caption = "Opportunity-adjusted attempts with 95% game-cluster intervals")

net_re_intervals <- team_intervals[metric == "net_re"]
net_re_table <- merge(
  teams_display[, .(team_id, team, net_re)],
  net_re_intervals[, .(team_id, lower, upper)],
  by = "team_id"
)[order(-net_re)]
net_re_display <- net_re_table[seq_len(min(15L, nrow(net_re_table))), .(
  Team = team, `Net RE` = net_re, Lower = lower, Upper = upper
)]
kable(net_re_display, digits = 3,
  caption = "Top 15 teams by net challenge RE with 95% game-cluster intervals")
```

# Sensitivity and limitations

The boundary sensitivity removes pitches within 0.25 inches of the official
ABS edge. This targets public-tracking and rounding uncertainty without
reclassifying the remaining pitches.

```{r}
sensitivity <- ledger[, .(
  correctable_all = sum(correctable_opportunity, na.rm = TRUE),
  correctable_outside_boundary = sum(
    correctable_opportunity & !boundary_sensitivity_excluded, na.rm = TRUE),
  out_all = sum(opportunity_status == "out_of_challenges", na.rm = TRUE),
  out_outside_boundary = sum(
    opportunity_status == "out_of_challenges" & !boundary_sensitivity_excluded,
    na.rm = TRUE)
)]
kable(sensitivity)
```

Unchallenged ABS outcomes remain geometry-based inferences. The RE288 and
home-win GAM are empirical 2023–2025 models, and all adjusted comparisons may
retain unmeasured differences in communication, pitcher/catcher confidence,
and game context.
