---
title: "Revealed Selection and Perception-Aware ABS Challenge Policy"
subtitle: "Five-fold cross-validated exploratory analysis"
author: "JP fork research pipeline"
date: last-modified
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(arrow)
library(data.table)
library(ggplot2)
library(jsonlite)
library(knitr)
library(rprojroot)

root <- find_root(has_file("_targets.R"), path = getwd())
derived <- file.path(
  root, "data", "processed", "perception", "revealed_policy"
)
artifact_names <- c(
  selection = "challenge_selection_oof.parquet",
  density = "challenge_margin_distributions.parquet",
  profile = "perception_profile.parquet",
  belief = "subjective_belief_envelope.parquet",
  thresholds = "policy_thresholds.parquet",
  value = "policy_value_summary.csv",
  manifest = "model_manifest.json"
)
artifact_paths <- stats::setNames(
  file.path(derived, unname(artifact_names)), names(artifact_names)
)
if (!all(file.exists(artifact_paths))) {
  stop(
    "Revealed-policy artifacts are missing. Run ",
    "`make revealed-policy-full` before rendering this report."
  )
}

selection <- as.data.table(read_parquet(artifact_paths[["selection"]]))
density <- as.data.table(read_parquet(artifact_paths[["density"]]))
profile <- as.data.table(read_parquet(artifact_paths[["profile"]]))
belief <- as.data.table(read_parquet(artifact_paths[["belief"]]))
thresholds <- as.data.table(read_parquet(artifact_paths[["thresholds"]]))
policy_value <- fread(artifact_paths[["value"]])
manifest <- read_json(artifact_paths[["manifest"]], simplifyVector = TRUE)
for (column in c(
  "gain_over_observed_re_lower_95", "gain_over_observed_re_upper_95"
)) {
  if (!column %in% names(policy_value)) policy_value[, (column) := NA_real_]
}

diagnostic_path <- function(name) file.path(derived, paste0(name, ".csv"))
read_diagnostic <- function(name) {
  path <- diagnostic_path(name)
  if (file.exists(path)) fread(path) else data.table()
}
selection_gate <- read_diagnostic("selection_gate")
selection_validation <- read_diagnostic("selection_validation")
official_validation <- read_diagnostic("policy_official_validation")
prior_scores <- read_diagnostic("prior_oof_log_scores")

snapshot_columns <- lapply(
  list(selection, density, profile, belief, thresholds, policy_value),
  function(x) unique(x$snapshot_id)
)
if (any(lengths(snapshot_columns) != 1L) ||
    length(unique(unlist(snapshot_columns))) != 1L) {
  stop("Primary artifacts do not share exactly one snapshot identifier")
}

log_loss <- function(y, p) {
  p <- pmin(1 - 1e-12, pmax(1e-12, p))
  mean(-(y * log(p) + (1 - y) * log1p(-p)))
}
finite_min <- function(x) if (any(is.finite(x))) min(x, na.rm = TRUE) else NA_real_
finite_max <- function(x) if (any(is.finite(x))) max(x, na.rm = TRUE) else NA_real_
```

# What is observed, and what is inferred

The primary quantity is directly tied to observed behavior:

\[
\pi_a(M,z)=P(\text{challenge}\mid M,z,\text{inventory}>0),
\]

where the role-oriented true margin (M) is positive when the initial call was
objectively wrong. For offense, this is the physical signed distance; for
defense, its sign is reversed. Every inventory-available adverse call is an
opportunity, whether challenged or passed.

Four quantities must remain distinct:

1. **Objective wrongness:** (M>0), from held-out tracked geometry.
2. **Challenge propensity:** the cross-fitted probability that the human unit
   challenges at a given true margin and context.
3. **Model-implied subjective belief:** (q(r)=P(M>0\mid R=r,z)), conditional
   on an assumed sensory share of the effective behavioral width.
4. **Normative policy:** challenge when (q(r)>q^*=L/(G+L)), where (G) is
   the immediate run-expectancy correction and (L) is the Bellman continuation
   value of the inventory put at risk.

No tracked margin is exposed to the replayed decision rule. It enters only to
integrate action probabilities over possible private signals and to evaluate
counterfactual success. Official overturns are joined only after all fits and
policies are frozen.

# Revealed challenge selection

```{r}
selection_summary <- selection[, .(
  opportunities = .N,
  challenges = sum(action),
  observed_rate = mean(action),
  predicted_rate = mean(oof_propensity),
  heldout_log_loss = log_loss(action, oof_propensity),
  unit_fallback_rate = mean(unit_fallback),
  team_fallback_rate = mean(team_fallback),
  selected_model = paste(sort(unique(selected_model)), collapse = ", ")
), by = role]
kable(
  selection_summary, digits = 4,
  caption = "Five-fold out-of-fold challenge/pass prediction"
)

selection[, margin_bin := round(true_margin_inches * 2) / 2]
selection_curve <- selection[
  margin_bin >= -8 & margin_bin <= 8,
  .(
    opportunities = .N,
    observed = mean(action),
    fitted = mean(oof_propensity)
  ),
  by = .(role, tail_speed = fifelse(
    speed_stratum %in% c("slow_non_eephus", "eephus"),
    "slow / eephus", "ordinary or missing speed"
  ), margin_bin)
]
ggplot(selection_curve, aes(margin_bin, fitted, color = tail_speed)) +
  geom_vline(xintercept = 0, color = "grey65") +
  geom_line(linewidth = 0.9) +
  geom_point(aes(y = observed, size = opportunities), alpha = 0.35) +
  facet_wrap(~ role) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    x = "True role-oriented margin (inches; positive = call wrong)",
    y = "Challenge probability",
    color = "Speed stratum",
    size = "Opportunities",
    title = "Observed and fitted challenge selection over the full support"
  ) +
  theme_minimal()
```

The speed-aware full-support GAM is promoted separately by role only if its
held-out log loss improves by one game-clustered standard error, or if it fixes
a predeclared tail-calibration problem without a one-SE overall loss. Otherwise
the local common-probit baseline remains primary. Slow and eephus tails are
retained and diagnosed separately.

```{r}
if (nrow(selection_gate)) {
  kable(
    selection_gate, digits = 4,
    caption = "Predeclared full-support model promotion gate"
  )
}
```

# Opportunity priors and challenged margins

The role prior (f(M\mid z)) is fitted from all tracked adverse calls, including
rows with no available challenge. Shared Gaussian component shapes are combined
with count-by-coarse-pitch-family weights. Within each outer training fold,
(K\in\{1,3,6\}) is chosen by the clustered one-SE location-score rule; a
count-only weighting is retained as a sensitivity.

```{r}
if (nrow(prior_scores) && all(c(
  "location_log_density", "count_only_log_density"
) %in% names(prior_scores))) {
  prior_sensitivity <- prior_scores[, .(
    heldout_pitches = .N,
    count_x_family_mean_log_density = mean(location_log_density),
    count_only_mean_log_density = mean(count_only_log_density),
    count_x_family_minus_count_only = mean(
      location_log_density - count_only_log_density
    )
  ), by = role]
  kable(
    prior_sensitivity, digits = 4,
    caption = "Count-by-family prior versus count-only sensitivity"
  )
}
```

The challenged-margin distribution is derived, not separately fitted:

\[
f(M\mid A=1,z)=
\frac{f(M\mid z)\pi_a(M,z)}
{\int f(u\mid z)\pi_a(u,z)\,du}.
\]

```{r}
density_near <- density[margin_inches >= -8 & margin_inches <= 8]
density_plot <- density_near[, .(
  opportunity_density = weighted.mean(
    opportunity_density, heldout_context_rows
  ),
  challenged_density = weighted.mean(
    selected_density, heldout_context_rows * selected_action_mass
  )
), by = .(role, margin_inches)]
density_plot <- melt(
  density_plot,
  id.vars = c("role", "margin_inches"),
  variable.name = "distribution", value.name = "density"
)
ggplot(density_plot, aes(margin_inches, density, color = distribution)) +
  geom_vline(xintercept = 0, color = "grey65") +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ role, scales = "free_y") +
  labs(
    x = "Role-oriented margin (inches; positive = call wrong)",
    y = "Density",
    color = NULL,
    title = "Opportunity and model-implied challenged-margin densities"
  ) +
  theme_minimal()

wrong_mass <- unique(density[, .(
  fold, role, prior_context,
  opportunity_wrong_call_mass_analytic,
  selected_wrong_call_mass,
  selected_action_mass,
  heldout_context_rows
)])[, .(
  opportunity_wrong_call_rate = weighted.mean(
    opportunity_wrong_call_mass_analytic, heldout_context_rows
  ),
  model_implied_challenge_success_rate = weighted.mean(
    selected_wrong_call_mass,
    heldout_context_rows * selected_action_mass
  )
), by = role]
kable(
  wrong_mass, digits = 4,
  caption = "Prior wrong-call mass and propensity-selected success mass"
)
```

# What perception is—and is not—identified

The local challenge curve identifies a common effective behavioral width
(w_a). It does not uniquely separate sensory uncertainty from action friction.
The report therefore profiles

\[
\sigma_{\text{sensory},a}=\kappa_a w_a,\qquad
\tau_{\text{action},a}=w_a\sqrt{1-\kappa_a^2},
\]

for all 25 offense/defense combinations with
(\kappa_a\in\{0,.25,.5,.75,1\}). A scenario belongs to the identified set if
its joint held-out challenge-choice loss is within one paired,
game-clustered standard error of the best scenario.

```{r}
scenario_profile <- unique(profile[, .(
  scenario_id, offense_kappa, defense_kappa,
  joint_heldout_log_loss, difference_from_best,
  one_se_membership, behaviorally_supported, interpretation
)])
ggplot(
  scenario_profile,
  aes(factor(offense_kappa), factor(defense_kappa),
      fill = joint_heldout_log_loss)
) +
  geom_tile(color = "white") +
  geom_point(
    data = scenario_profile[one_se_membership == TRUE],
    shape = 21, fill = NA, color = "black", size = 4, stroke = 1
  ) +
  labs(
    x = "Offense sensory share (kappa)",
    y = "Defense sensory share (kappa)",
    fill = "Held-out\nlog loss",
    title = "Profiled sensory/action decompositions",
    subtitle = "Outlined cells are within one game-clustered SE of the best"
  ) +
  theme_minimal()

widths <- unique(profile[, .(
  fold, role, effective_width_inches, kappa,
  sensory_sigma_inches, action_sigma_inches,
  one_se_membership, interpretation
)])
kable(
  widths[one_se_membership == TRUE][order(fold, role, kappa)],
  digits = 3,
  caption = "Effective, sensory, and action widths in the identified set"
)

belief_all <- belief[
  count_state == "all_counts" &
    perceived_margin_inches >= -8 & perceived_margin_inches <= 8
]
ggplot(
  belief_all,
  aes(perceived_margin_inches, subjective_q_mean)
) +
  geom_ribbon(
    aes(ymin = subjective_q_min, ymax = subjective_q_max),
    fill = "#4C78A8", alpha = 0.2
  ) +
  geom_line(color = "#2457A6", linewidth = 0.9) +
  geom_vline(xintercept = 0, color = "grey65") +
  facet_wrap(~ role) +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
  labs(
    x = "Perceived role-oriented margin r (inches)",
    y = "Model-implied P(call wrong | r)",
    title = "Subjective-belief envelope over the identified set",
    subtitle = paste(
      "Band spans accepted scenarios, outer folds, and contextual",
      "pitch-family priors; it is not a confidence interval"
    )
  ) +
  theme_minimal()
```

```{r}
if (!all(scenario_profile$behaviorally_supported %in% TRUE)) {
  cat(
    "**The structural profile is not non-inferior to the direct selection ",
    "benchmark by the predeclared one-SE rule. Subjective-belief curves and ",
    "policy values must therefore be read as assumption-driven sensitivity ",
    "analyses, not recovered player beliefs.**\n"
  )
}
```

# Bellman thresholds and shared inventory

Each held-out opportunity receives its immediate correction gain (G), the
training-fold Bellman continuation loss (L), and the payoff threshold
(q^*=L/(G+L)). Offense and defense share the same chronological team inventory;
failed challenges consume one, successful challenges retain it, and no future
tracked truth enters the action rule.

```{r}
threshold_summary <- thresholds[, .(
  opportunities = sum(opportunity_rows),
  mean_G = weighted.mean(G, opportunity_rows),
  mean_L = weighted.mean(L, opportunity_rows),
  mean_q_star = weighted.mean(q_star, opportunity_rows),
  finite_signal_threshold_rate = weighted.mean(
    is.finite(model_implied_r_star_inches), opportunity_rows
  ),
  context_fallback_rate = sum(context_fallback_rows) / sum(opportunity_rows),
  numerical_root_fallback_rate =
    sum(numerical_root_fallback_rows) / sum(opportunity_rows)
), by = .(role, inventory)]
kable(
  threshold_summary, digits = 4,
  caption = "Normative payoff and signal thresholds across accepted scenarios"
)
```

# Expected-run policy comparison

The held-out replay compares observed choices, the fitted human-selection
policy, every accepted normative perception scenario, no challenges, and the
exact-location oracle. Run expectancy is the objective; WPA is not used.

```{r}
combined <- policy_value[role == "combined"]
comparators <- combined[!grepl("^normative__", policy)]
kable(
  comparators[, .(
    policy, policy_family, captured_re, attempts, success_rate,
    inventory_exhaustion_rate, gain_over_observed_re,
    gain_over_observed_re_lower_95, gain_over_observed_re_upper_95
  )],
  digits = 3,
  caption = "Combined offense/defense expected-run comparison"
)

normative <- policy_value[grepl("^normative__", policy)]
normative_envelope <- normative[, .(
  scenarios = .N,
  gain_lower = min(gain_over_observed_re),
  gain_upper = max(gain_over_observed_re),
  attempts_lower = min(attempts),
  attempts_upper = max(attempts),
  success_rate_lower = min(success_rate),
  success_rate_upper = max(success_rate),
  bootstrap_lower = finite_min(gain_over_observed_re_lower_95),
  bootstrap_upper = finite_max(gain_over_observed_re_upper_95)
), by = role]
kable(
  normative_envelope, digits = 3,
  caption = "Normative policy envelope over the identified set"
)
```

# Final official-outcome validation

Official outcomes never train the prior, selection model, perception profile,
or Bellman policy. They enter here only to compare official overturns with the
geometry-implied result among observed challenges. Near-perfect agreement is a
tracking/label validation and is not evidence that perceived location was
observed.

```{r}
if (nrow(official_validation)) {
  kable(
    official_validation, digits = 4,
    caption = "Official versus geometry challenge-result validation"
  )
}
if (nrow(selection_validation)) {
  kable(
    selection_validation, digits = 4,
    caption = "Held-out revealed-selection diagnostics"
  )
}
```

# Scope

These estimates use all games in the current target snapshot with five-fold
game separation. They are cross-validated and exploratory because no untouched
temporal confirmation set remains. “Perception-aware” means that policy values
are reported across plausible latent sensory channels. It does not mean that a
player's private perceived pitch location, sensory width, or subjective success
probability was directly observed or point-identified.

```{r}
cat("Snapshot:", unique(selection$snapshot_id), "\n\n")
cat("Git SHA:", manifest$git_sha, "\n\n")
cat("Profile:", manifest$profile, "\n\n")
cat("Games:", manifest$games, "\n")
```
